---
title: "Data Quality Assignment"
author: "Gilbert Habaasa"
date: "2025-04-04"
output:
  pdf_document: default
  html_document: default
---

# IPUMS Exercise

## IPUMS International Data Extract and Analysis

### Data Quality, European Doctoral School of Demography 2024-2025

### INED – Paris (France)

### Instructor: Mariona Lozano, Centre d’Estudis Demogràfics

### Student: Gilbert Habaasa


### **Getting IPUMS Data into R**

```{r}
options(repos = c(CRAN = "https://cloud.r-project.org"))
install.packages("ipumsr")
```

```{r}
setwd("C:\\Users\\admin\\OneDrive - London School of Hygiene and Tropical Medicine\\‌INED 2024\\Data Quality\\IPUMS")
# file.exists("C:\\Users\\admin\\OneDrive - London School of Hygiene and Tropical Medicine\\‌INED 2024\\Data Quality\\IPUMS\\ipumsi_00001.dat.gz")

library(ipumsr) 
library(tidyverse)
library(readr)
ddi <- read_ipums_ddi("ipumsi_00001.xml")
data <- read_ipums_micro(ddi) 
data
```

### **Question and Variables**

In this exercise you will gain basic familiarity with the IPUMS International data exploration and extract system to 
answer the following question: “What are the differences in urbanization, literacy, and occupational participation in 
Uganda and Mexico?” You will create a data extract that includes the following variables: **URBAN, SEX, EMPSTAT, OCCISCO, 
LIT, AGE**. 

**Variables and Code**

URBAN: household location
  1 = Rural
  2 = Urban
  9 = Unknown

SEX 
  1 = Male
  2 = Female
  9 = Missing/blank

URBAN: EMPSTAT: Employment status 
  1 = Employed
  2 = Unemployed
  3 = Not in labor force
  9 = Unknown/Illegible

OCCISCO: Employment category 
  01 = Legislators, senior officials and managers
  02 = Professionals
  03 = Technicians and associate professionals
  04 = Clerks
  05 = Service workers and shop and market sales
  06 = Skilled agricultural and fishery workers
  07 = Crafts and related trades workers
  08 = Plant and machine operators and assemblers
  09 = Elementary occupations
  10 = Armed forces
  11 = Other occupations, unspecified or n.e.c.
  97 = Response suppressed
  98 = Unknown
  99 = NIU (not in universe)

LIT: Literacy 
  0 = NIU (not in universe)
  1 = No, illiterate
  2 = Yes, literate	
  9 = Unknown, illegible or blank

 AGE 
  000 = Less than 1 year old
  001 = 1
  ... = ...
  140 = 140
  999 = Missing

 COUNTRY 
  484 = Mexico
  800 = Uganda



### **Analyse the Data**

### **Part 1: Variable documentation**
For each variable below, search through the tabbed sections of the variable description to answer each question.

1.Under the “Household” dropdown menu, find the “Geography” subcategory and click on the variable URBAN. What constitutes an urban area in each country?
a.Mexico 2000:

**Urban places are defined consistently across Mexican samples as localities with 2,500 or more persons.**

b.Uganda 2002:

**Urban areas in 2002 and 2014 are gazetted cities, municipalities and towns with more than 2,000 inhabitants.**

2.What are the codes for URBAN?

**1.Rural; 2.Urban; 9.Unknown.**

3.Find the variable EMPSTAT. Is the reference period of work the same for Mexico and Uganda?

**Mexico-Last week**

**Uganda-Last seven days**

4.What is the universe for EMPSTAT in: 

a.Mexico 2000?

**Mexico-Persons age 12+**

b.Uganda 2002?

**Uganda-Persons age 5+**


### **Part 2. Frequencies**

5.Find codes page for the SAMPLE variable. What are the codes for: 

a.Mexico 2000? 

**Value Code for Mexico-484**

b.Uganda 2002?

**Value Code for Uganda-800**

```{r}
unique(data$COUNTRY)
```

6.How many individuals are in the Mexico 2000 sample extract?

**There are 10099182 individuals in the Mexico 2000 sample extract**

```{r}
data |> filter(COUNTRY==484) |> count()
```


7.How many individuals are in the Uganda 2002 sample extract?

**There are 2497449	individuals in the Uganda 2002 sample extract**

```{r}
data |> filter(COUNTRY==800) |> count()
```


8.What proportion of individuals in the sample lived in urban areas in each country?

a.Mexico 2000:

**In Mexico, 59.2% of individuals in the 2000 sample lived in urban areas.**

b.Uganda 2002:

**In Uganda, 12.3% of individuals in the 2002 sample lived in urban areas.**

```{r}
100*prop.table(table(data$URBAN,data$COUNTRY),2)
```



### **Part 3. Weighted frequencies**

To get a more accurate estimate of the actual proportion of individuals living in urban areas, you will have to use the person weight. 

9.Using weights, what is the total population of each country?

a.Mexico 2000:

**97014867 people**

b.Uganda 2002:

**24974490 people**

```{r}
data |> filter(COUNTRY==484) |> summarise(total_population=sum(PERWT))
```
```{r}
data |> filter(COUNTRY==800) |> summarise(total_population=sum(PERWT))
```


10.Using weights, how many individuals lived in urban areas in each country?

a.Mexico 2000:

**72409464 people**

b.Uganda 2002:

**3060540 people**

```{r}
data |> group_by (COUNTRY, URBAN) %>% summarise(living_in_urban=sum(PERWT))
```

11.Using weights, what proportion of individuals lived in urban areas in each country?

a.Mexico: 2000:

**Using weights, 74.6% of individuals in the Mexico 2000 sample lived in urban areas.**

b.Uganda 2002:

**Using weights, 12.3% of individuals in the Uganda 2002 sample lived in urban areas.**

```{r}
data |> 
  group_by(COUNTRY) |> 
  summarise(prop_urban = sum(PERWT[URBAN == 2]) / sum(PERWT) * 100)
```


### **Part 4. Trends**

12.Using weights, which occupational category has the highest percentage of workers?

a.In Mexico 2000:

***Crafts and related trades workers (17.9%)***


b.In Uganda 2002:

***Skilled agricultural and fishery workers (71.4%)***


```{r}
data |> filter(OCCISCO != 98 & OCCISCO != 99 ) |> 
  group_by(COUNTRY, OCCISCO) |> 
  summarise(weighted_n = sum(PERWT, na.rm = TRUE), .groups = "drop") |> 
  group_by(COUNTRY) |> 
  mutate(perc = weighted_n / sum(weighted_n) * 100)|> 
  arrange(desc(perc))  |> print(n=50)
```


13.Which occupation category has the highest percentage of female workers?

a.In Mexico 2000:


***Code # 05=Service workers and shop and market sales (7.7%)***


b.In Uganda 2002:

***Code # 06 = Skilled agricultural and fishery workers (35.3%)***

```{r}
data |> filter(OCCISCO != 98 & OCCISCO != 99) |> 
  group_by(COUNTRY, OCCISCO, SEX) |> 
  summarise(weighted_n = sum(PERWT, na.rm = TRUE), .groups = "drop") |> 
  group_by(COUNTRY) |> 
  mutate(perc = weighted_n / sum(weighted_n) * 100)|> 
  arrange(desc(perc))  |> print(n=50)
```



14.What is the labour force participation distribution by gender in each country?

a.Mexico 2000:

***Men-71.7% ; Female-31.4%***

b.Uganda 2002: 

***Men-43.5% ; Female-33.7%***


```{r}
unique(data$EMPSTAT)
data |> 
  filter(EMPSTAT %in% c(1, 2, 3)) |>  # esclude 0, 9
  mutate(
    labour_force = ifelse(EMPSTAT %in% c(1, 2), 1, 0)
  ) |> 
  group_by(COUNTRY, SEX) |> 
  summarise(
    total = sum(PERWT, na.rm = TRUE),
    lf = sum(PERWT * labour_force, na.rm = TRUE),
    lfpr = lf / total * 100,
    .groups = "drop"
  )
```

15.What percentage of women within the labour force is working:

a.In agriculture in Mexico 2000:

 ***4.7% of women within labourforce in Mexico work in agriculture***

b.In agriculture in Uganda 2002:

 ***76.5% of women within labourforce in Uganda work in agriculture***

c.In service in Mexico 2000: 

 ***23.7% of women within labourforce in Uganda work in services employment***

d.In service in Uganda 2002:

 ***8.7% of women within labourforce in Uganda work in services employment***
 

```{r}
unique(data$OCCISCO)
```


```{r}

# AGRICULTURE IN MEXICO AND UGANDA

# MEXICO
# Women labour force in Mexico 
subset_data <- data[data$COUNTRY == 484 & 
                      data$SEX == 2 & 
                      data$EMPSTAT %in% c(1, 2), ]

# Occupational categories
agriculture_codes <- c("6")  # Agricultural workers

# Numerator 1: Women in agriculture occupation (OCCISCO == 6)
numerator1 <- sum(subset_data$PERWT[subset_data$EMPSTAT == 1 & subset_data$OCCISCO == 6], na.rm = TRUE)

# Denominator: Total women in labour force (working+Non-working)
denominator1 <- sum(subset_data$PERWT, na.rm = TRUE)


# Mexico Percent of Women in labourforce working in agriculture
mexico_perc_agric <- (numerator1 / denominator1) * 100
mexico_perc_agric

#UGANDA
# Women labour force in Uganda 
subset_data <- data[data$COUNTRY == 800 & 
                      data$SEX == 2 & 
                      data$EMPSTAT %in% c(1, 2), ]

# Numerator 2: Women in agriculture occupation (OCCISCO == 6)
numerator2 <- sum(subset_data$PERWT[subset_data$EMPSTAT == 1 & subset_data$OCCISCO == 6], na.rm = TRUE)

# Denominator 2: Total women in labour force (working+Non-working)
denominator2 <- sum(subset_data$PERWT, na.rm = TRUE)


# Uganda Percent of Women in labourforce working in agriculture
uganda_perc_agric <- (numerator2 / denominator2) * 100
uganda_perc_agric

```

```{r}

# SERVICES IN MEXICO AND UGANDA

# MEXICO
# Women labour force in Mexico 
subset_data <- data[data$COUNTRY == 484 & 
                      data$SEX == 2 & 
                      data$EMPSTAT %in% c(1, 2), ]

# Occupational categories
service_codes <- c("5")  # Service workers

# Numerator 3: Women in service occupation (OCCISCO == 5)
numerator3 <- sum(subset_data$PERWT[subset_data$EMPSTAT == 1 & subset_data$OCCISCO == 5], na.rm = TRUE)

# Denominator 3: Total women in labour force (working+Non-working)
denominator3 <- sum(subset_data$PERWT, na.rm = TRUE)


# Mexico Percent of Women in labourforce working in service Occupation 
mexico_perc_services <- (numerator3 / denominator3) * 100
mexico_perc_services

#UGANDA
# Women labour force in Uganda 
subset_data <- data[data$COUNTRY == 800 & 
                      data$SEX == 2 & 
                      data$EMPSTAT %in% c(1, 2), ]

# Numerator 4: Women in Service occupation (OCCISCO == 6)
numerator4 <- sum(subset_data$PERWT[subset_data$EMPSTAT == 1 & subset_data$OCCISCO == 5], na.rm = TRUE)

# Denominator 4: Total women in labour force (working+Non-working)
denominator4 <- sum(subset_data$PERWT, na.rm = TRUE)


# Uganda Percent of Women in labourforce working in Service Occupation
uganda_perc_services <- (numerator4 / denominator4) * 100
uganda_perc_services

```

### **Part 5: Graphical Analysis**

16.What percentage of the population is literate in each country?

a.Mexico 2000:

***77.7% of the Population in Mexico 2000 is literate***

b.Uganda 2002:

***45.1% of the Population in Uganda 2002 is literate***


```{r}
#Mexico 2000: Percentage literate, excluding response suppressed, unknown, and NIU
mexico_literacy <- data %>%
  filter(COUNTRY == 484, YEAR == 2000, LIT == 2, !LIT %in% c(97, 98, 99)) %>%
  summarise(literate_weight = sum(PERWT, na.rm = TRUE)) %>%
  mutate(
    mexico_literacy_percentage = literate_weight / 
      sum(data$PERWT[data$COUNTRY == 484 & data$YEAR == 2000], na.rm = TRUE) * 100
  )
mexico_literacy$mexico_literacy_percentage

#Uganda 2002: Percentage literate, excluding response suppressed, unknown, and NIU
uganda_literacy <- data %>%
  filter(COUNTRY == 800, YEAR == 2002, LIT == 2, !LIT %in% c(97, 98, 99)) %>%
  summarise(literate_weight = sum(PERWT, na.rm = TRUE)) %>%
  mutate(
    uganda_literacy_percentage = literate_weight / 
      sum(data$PERWT[data$COUNTRY == 800 & data$YEAR == 2002], na.rm = TRUE) * 100
  )

uganda_literacy$uganda_literacy_percentage
```

17.(OPTIONAL) Create a graph to visualize differences in the percentage of literacy by AGE and SEX in both countries. 

```{r}
data |> 
  filter(LIT %in% 1:2, AGE < 100) |> 
  summarise(n = sum(PERWT), .by = c(COUNTRY, LIT, AGE)) |> 
  mutate(prop = round(100*n/sum(n),2), .by = c(COUNTRY,AGE)) |>
  mutate(
    COUNTRY = factor(COUNTRY, levels = c(484, 800), labels = c("Mexico","Uganda"))
  ) |> 
  filter(LIT == 2, AGE >= 10) |> 
  ggplot() +
  aes(x = AGE, y = prop, color = COUNTRY, group = COUNTRY) +
  geom_point(size = 2, alpha = .5, show.legend = FALSE) +
  geom_smooth(se = FALSE,linewidth = 1.3, show.legend = FALSE) +
  scale_color_manual(values = c("blue","red")) +
  scale_y_continuous(breaks = seq(0,100,20)) +
  scale_x_continuous(breaks = seq(0,100,10)) +
  labs(
    x = "Age (years)",
    y = "Proportion of literate population (%)",
    caption = "Source: IPUMS International Census data",
    title = "Proportion of literate population by age (%) in Mexico and Uganda"
  ) +
  annotate(
    geom="text",
    x=80,
    y=80,
    label="Mexico 2000",
    color="black",
    fontface =2,
    size = 6
  ) +
  annotate(
    geom="text",
    x=50,
    y=30,
    label="Uganda 2002",
    color="black",
    fontface =2,
    size = 6
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(color = "black", face = "bold", hjust = .5),
    plot.caption = element_text(color = "black", face = "italic", hjust = 1),
    axis.title.x = element_text(color = "black", face = "bold", hjust = .5),
    axis.title.y = element_text(color = "black", face = "bold", hjust = .5),
    axis.text = element_text(color = "black")
  )
```





